Model complexity and model choice for animal movement models

### 2016-06-08 guelph_talk/guelph_panther.Rnw
guelph talk ...

\usetheme{Frankfurt}
\usecolortheme{dove}
\setbeamercovered{transparent}
\setbeamercolor{description item}{fg=blue}
\newcommand{\citepx}[1]{{\small \citep{#1}}}
\newcommand{\citex}[1]{{\small \cite{#1}}}
% \renewcommand{\descriptionlabel}[1]{\hspace{\labelsep}\textbf{#1}}
\usepackage{babel}

@@

\begin{frame}
\frametitle{Acknowledgements}
\begin{description}
\item[People] Michael Li, Madelon van de Kerk, Dave Onorato
\item[Support] NSERC Discovery grant
\end{description}

@@

\begin{itemize}
\item observations:  \\
e.g. mass mark-recapture, longitudinal density,
direct observation, telemetry (VHF, GPS)
\item most methods provide a sequence of times and locations for each individual
\item summaries:
\begin{itemize}
\item home range \\
(convex hull, kernel density estimate, etc.)
\item root-mean-squared displacement
\item step length and turning angle
\end{itemize}
\item covariates: \\
e.g. habitat map, individual characteristics (sex, age, weight \ldots)
\end{itemize}

@@

\begin{columns}
  \begin{column}{0.5\textwidth}
    \begin{itemize}
    \item Florida panther: \emph{Felix ...}
    \item severely reduced habitat, endangered
    \item small, isolated population
    \item currently recovering
    \end{itemize}
  \end{column}
\end{columns}

@@

\begin{columns}
  \begin{column}{0.5\textwidth}
    \begin{itemize}
    \item abc
    \end{itemize}
  \end{column}
\end{columns}

@@

\begin{frame}
\frametitle{example movement tracks}
FIXME: picture here!
(animation?? map background??)
\end{frame}

@@

\begin{itemize}
\item finite mixture model with temporal dependence
\item discrete time steps
\item discrete latent state; \emph{transition matrix}
\item observations from \emph{emission distributions} \\
(continuous or discrete, univariate or multivariate)
\end{itemize}

@@

@@

\begin{itemize}
\item \emph{forward-backward algorithm} for estimating parameters
\item \emph{Viterbi algorithm} for estimating most probable state sequences
\item {\tt depmixS4} package \citepx{visser2010depmixs4}
(also {\tt moveHMM} \citepx{michelot_movehmm_2016}, {\tt move.HMM} \citepx{augustine_hmm})
\end{itemize}

@@

\section[Basic analysis]{Basic analysis (vdK et al.)}
\subsection{}

@@

@@

\includegraphics[height=\textheight]{pix/jae_fig1.png}

>>>>

\newcommand{\fixme}[1]{{\color{red} \textbf{#1}}}
\usetheme{Frankfurt}
\usecolortheme{dove}
\setbeamercovered{transparent}
\setbeamercolor{description item}{fg=blue}
\newcommand{\citepx}[1]{{\small \citep{#1}}}
\newcommand{\citex}[1]{{\small \cite{#1}}}
% \renewcommand{\descriptionlabel}[1]{\hspace{\labelsep}\textbf{#1}}
\usepackage{babel}

@@

\begin{frame}
\frametitle{Acknowledgements}
\begin{description}
\item[People] \textbf{Michael Li}, \textbf{Madelon van de Kerk}, Dave Onorato, Madan Oli
\item[Agencies] US Fish and Wildlife, US Geological Survey, US National Park Service
\item[Funding] NSERC Discovery grant, NSF IGERT program
\end{description}

@@

\begin{itemize}
\item observations:  \\
e.g. mass mark-recapture, longitudinal density,
direct observation, telemetry (VHF, GPS)
\item most methods provide a sequence of times and locations for each individual
\item summaries:
\begin{itemize}
\item home range \\
(convex hull, kernel density estimate, etc.)
\item root-mean-squared displacement
\item step length and turning angle
\end{itemize}
\item covariates: \\
e.g. habitat map,  \\
individual characteristics (sex, age, weight \ldots)
\end{itemize}

@@

\begin{columns}
  \begin{column}{0.5\textwidth}
    \begin{itemize}
    \item Florida panther: \emph{Felix ...}
    \item severely reduced habitat, endangered
    \item small, isolated population
    \item currently recovering
    \end{itemize}
  \end{column}
  \begin{column}{0.5\textwidth}
    (cute panther pictures here)
  \end{column}
\end{columns}

@@

\begin{columns}
  \begin{column}{0.5\textwidth}
    \begin{itemize}
    \item how does panther movement vary by sex and life history stage \\
      (juvenile, adult, mom with kittens \ldots)
    \item how does movement mediate exposure to threats \\
      (intraspecific aggression, roadkill) ?
    \item can we predict the effects of future changes in 
      population density / population structure / habitat?
    \end{itemize}
  \end{column}
  \begin{column}{0.5\textwidth}
    (more cute pictures here ... or roadkill?)
  \end{column}
\end{columns}

@@

\begin{frame}
\frametitle{example movement tracks}
\includegraphics[width=\textwidth]{pix/track1.pdf}
\end{frame}

@@

\begin{itemize}
\item finite mixture model with temporal dependence
\item discrete time steps
\item discrete latent state; \emph{transition matrix}
\item observations from \emph{emission distributions} \\
(continuous or discrete, univariate or multivariate)
\item \textbf{multiphasic movement} \citepx{fryxell_multiple_2008,langrock_flexible_2012}
\end{itemize}

@@

\newcommand{\obs}{\mathbf{y}}
\newcommand{\cov}{\mathbf{x}}
\newcommand{\state}{z}

@@

\fixme{notation, underbraces}

\begin{equation}
\begin{split}
P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T}|\boldsymbol \theta,\mathbf{X}_{1:T}) & =
P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \cdot \\
&  \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
\end{split}
\label{eq:HMMlik}
\end{equation}

@@

\begin{frame}
\frametitle{Hidden Markov models (part 3)}

\begin{itemize}
\item \emph{forward-backward algorithm} for estimating parameters
\item \emph{Viterbi algorithm} for estimating most probable state sequences
\item {\tt depmixS4} package \citepx{visser2010depmixs4}
(also {\tt moveHMM} \citepx{michelot_movehmm_2016})
\item hidden \emph{semi-Markov} models: allow for non-geometric \emph{dwell distributions} \citepx{langrock_hidden_2011-1,augustine_hmm}: {\tt move.HMM}
\end{itemize}

\section[Basic analysis]{Basic analysis \citepx{kerk2015hidden}}
\subsection{}

@@

\begin{frame}
\frametitle{Tracks with Viterbi estimates}

\includegraphics[width=\textwidth]{pix/track1C.pdf}

@@

\includegraphics[height=0.8\textheight]{pix/BICplot.pdf}
